---
title: "F1 Project Data Visualization"
author: "Cole Wagner"
format:
  html:
    toc: true
    self-contained: true
    code-fold: true
    code-line-numbers: true
    code-tools: true
    cache: true
    theme: "yeti"
editor: visual
---

```{r setup}
library(tidyverse)
library(RColorBrewer)
library(plotly)
library(reactable)
library(htmltools)
f1_laps <- read.csv(
  "../Data/full_no_missing.csv"
)
```

## Boxplot for Tracks with Rain

```{r rain_box}
# Pull only tracks that had rain during their qualifying session
rainy_tracks <- unique(f1_laps$CIRCUIT_SHORT_NAME[f1_laps$RAINFALL == 1])

rain_box <- f1_laps %>%
  filter(CIRCUIT_SHORT_NAME %in% rainy_tracks) %>%
  plot_ly(
    type = "box"
  ) %>%
  # Add a trace for rainy laps
  add_trace(
    x = ~ CIRCUIT_SHORT_NAME[RAINFALL == 1],
    y = ~ LAP_DURATION[RAINFALL == 1],
    legendgroup = "Rain",
    name = "Rain",
    color = I("blue")
  ) %>%
  # Add a trace for dry laps
  add_trace(
    x = ~ CIRCUIT_SHORT_NAME[RAINFALL == 0],
    y = ~ LAP_DURATION[RAINFALL == 0],
    legendgroup = "No Rain",
    name = "No Rain"
  ) %>%
  # Set the axis and overall titles
  layout(
    title = "Distribution of Lap Times by Rain Status",
    yaxis = list(
      title = "Lap Duration (Seconds)"
    ),
    xaxis = list(
      title = "Circuit Name"
    ),
    boxmode = "group"
  )

rain_box
```

Why do Montreal, Hungaroring, and Spa all have lower average lap times when it is raining?

```{r rain_check}
# Filter to only see laps for the circuits in question
rain_checks <- f1_laps %>%
  filter(CIRCUIT_SHORT_NAME %in% c("Montreal", "Hungaroring", "Spa-Francorchamps")) %>%
  group_by(CIRCUIT_SHORT_NAME, YEAR) %>%
  summarise(
    total_laps = n(),
    laps_in_rain = sum(RAINFALL == 1)
  )

```

For Montreal, there was only rain in 13 of the 90 recorded laps. For Hungary, the rain wasn't severe enough to switch off standard soft tyres. For Spa, the rain was so heavy that everyone had to use intermediate tyres regardless of the current rain status.

## Overlapping Histograms for Lap Times by Circuit Type

```{r type_laptime}
type_laptime <- f1_laps %>%
  plot_ly(
    type = "histogram"
  ) %>%
  # Add a trace for race circuit laps
  add_trace(
    x = ~ LAP_DURATION[TYPE == "Race circuit"],
    legendgroup = "Race Circuit",
    name = "Race Circuit",
    color = I("dodgerblue"),
    alpha = 0.5,
    nbinsx = 20,
    histnorm = "percent",
    hovertemplate = paste(
      "Lap Time (sec): ", "%{x}",
      "<br>Percentage of Laps: ", "%{y}", "%"
    )
  ) %>%
  # Add a trace for street circuit laps
  add_trace(
    x = ~ LAP_DURATION[TYPE == "Street circuit"],
    legendgroup = "Street Circuit",
    name = "Street Circuit",
    color = I("firebrick1"),
    alpha = 0.5,
    nbinsx = 20,
    histnorm = "percent",
    hovertemplate = paste(
      "Lap Time (sec): ", "%{x}",
      "<br>Percentage of Laps: ", "%{y}", "%"
    )
  ) %>%
  # Adjust overall labels and format
  layout(
    barmode = "overlay",
    title = "Percentage Distribution of Lap Times by Circuit Type",
    xaxis = list(
      title = "Lap Time (Seconds)"
    ),
    yaxis = list(
      title = "Percentage of Laps"
    )
  )

type_laptime
```

# Bar Chart of Average Lap Time by Circuit

```{r circuit_bar}
circuit_times <- f1_laps %>%
  group_by(CIRCUIT_SHORT_NAME) %>%
  # Find mean and sd of lap times for each circuit
  summarise(
    avg_time = mean(LAP_DURATION),
    sd_time = sd(LAP_DURATION)
  ) %>%
  arrange(desc(avg_time))

circuit_bar <- ggplot(
  circuit_times,
  aes(
    # Plot the circuits in descending order of average lap time
    x = factor(CIRCUIT_SHORT_NAME,
      levels = CIRCUIT_SHORT_NAME
    ),
    y = avg_time,
    fill = CIRCUIT_SHORT_NAME
  )
) +
  geom_bar(stat = "identity", color = "black") +
  # Create 1 sd error bars
  geom_errorbar(
    ymin = circuit_times$avg_time - circuit_times$sd_time,
    ymax = circuit_times$avg_time + circuit_times$sd_time,
    width = 0.2
  ) +
  theme_bw() +
  theme(
    legend.position = "None",
    axis.text.x = element_text(angle = 45, hjust = 1, size = 14),
    axis.text.y = element_text(size = 14),
    title = element_text(size = 18, face = "bold"),
    axis.title = element_text(size = 16, face = "bold"),
    plot.caption = element_text(size = 8, face = "plain")
  ) +
  labs(title = "Average Lap Time for Each Circuit 2023-2024",
       x = "Circuit Name",
       y = "Average Lap Time",
       caption = "*Error bars represent one standard deviation.")

circuit_bar
```

```{r}
# Find the average of all weather data for each circuit in each year
weather_table <- f1_laps %>%
  group_by(CIRCUIT_SHORT_NAME, YEAR) %>%
  summarise(
    avg_temp = round(mean(AIR_TEMPERATURE), 3),
    avg_wind = round(mean(WIND_SPEED), 3),
    avg_hum = round(mean(HUMIDITY), 1),
    avg_pressure = round(mean(PRESSURE), 3),
    avg_rain = mean(RAINFALL),
    # Convert rain into a factor
    rain_factor = ifelse(avg_rain > 0, "Yes", "No")
  ) %>%
  select(-avg_rain)

temp_pal <- function(x) rgb(colorRamp(c("white", "firebrick1"))(x), maxColorValue = 255)

# Create a bar chart widget
bar_chart <- function(label, width = "100%", height = "1rem", fill = "#00bfc4", background = NULL) {
  bar <- div(style = list(background = fill, width = width, height = height))
  chart <- div(style = list(flexGrow = 1, marginLeft = "0.5rem", background = background), bar)
  div(style = list(display = "flex", alignItems = "center"), label, chart)
}

# Create a status badge widget
status_badge <- function(color = "#aaa", width = "0.55rem", height = width) {
  span(style = list(
    display = "inline-block",
    marginRight = "0.5rem",
    width = width,
    height = height,
    backgroundColor = color,
    borderRadius = "50%"
  ))
}

weather_reactable <- reactable(
  weather_table,
  columns = list(
    CIRCUIT_SHORT_NAME = colDef(name = "Circuit"),
    YEAR = colDef(name = "Year"),
    avg_temp = colDef(
      name = "Temperature (°C)",
      # Change cell background color based on temperature
      style = function(value) {
        normalized <- (value - min(weather_table$avg_temp)) / (max(weather_table$avg_temp) - min(weather_table$avg_temp))
        color <- temp_pal(normalized)
        list(background = color)
      }
    ),
    avg_wind = colDef(name = "Wind Speed (m/s)"),
    avg_hum = colDef(
      name = "Relative Humidity (%)",
      format = colFormat(percent = T, digits = 1),
      # Create bar chart based on relative humidity
      cell = function(value) {
        width <- paste0(value / max(weather_table$avg_hum) * 100, "%")
        bar_chart(value, width = width, fill = "skyblue", background = "#e1e1e1")
      }
    ),
    avg_pressure = colDef(name = "Pressure (mbar)"),
    rain_factor = colDef(
      name = "Rain",
      # Create a status badge based on whether it was raining or not
      cell = function(value) {
        color <- switch(value,
          Yes = "hsl(214, 45%, 50%)",
          No = "hsl(30, 97%, 70%)"
        )
        badge <- status_badge(color = color)
        tagList(badge, value)
      }
    )
  ),
  bordered = TRUE,
  highlight = TRUE,
  striped = TRUE,
  filterable = TRUE,
  defaultPageSize = 10,
  theme = reactableTheme(
    borderColor = "#959595",
    stripedColor = "ivory3",
    highlightColor = "#98F5FF",
    cellPadding = "8px 12px",
    searchInputStyle = list(width = "100%")
  )
)

weather_reactable

# saveRDS(weather_reactable, "../Data/weather_reactable.rds")
```
